SU($n$) ferromagnets near the Heisenberg critical point

### 2020-04-25 collective_sun/sun_perturbation.tex
overhauled lots of conventions in the SU(n) perturbation notes and shell model codes simplified the section on breaking SU(n) symmetry (perturbatively)

%%% symbols, notations, etc.
\usepackage{physics,braket,bm,amssymb} % physics and math
\renewcommand{\t}{\text} % text in math mode
\newcommand{\f}[2]{\dfrac{#1}{#2}} % shorthand for fractions
\newcommand{\p}[1]{\left(#1\right)} % parenthesis
\renewcommand{\sp}[1]{\left[#1\right]} % square parenthesis
\renewcommand{\set}[1]{\left\{#1\right\}} % curly parenthesis
\newcommand{\bk}{\Braket} % shorthand for braket notation

\renewcommand{\c}{\cdot} % inner product

\newcommand{\m}{\bm} % bold symbol
\renewcommand{\v}{\vec} % arrow vector

@@

\renewcommand{\d}{\text{d}}
\newcommand{\x}{\text{x}}
\newcommand{\y}{\text{y}}
\newcommand{\z}{\text{z}}

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\G}{\mathcal{G}}
\renewcommand{\H}{\mathcal{H}}
\newcommand{\I}{\mathcal{I}}
\renewcommand{\L}{\mathcal{L}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\T}{\mathcal{T}}

@@

\newcommand{\PS}{\text{PS}}
\newcommand{\EQPS}{=_{\text{PS}}}
\newcommand{\col}{\underline}

\let\var\relax
\DeclareMathOperator{\var}{var}
\DeclareMathOperator{\cov}{cov}
\DeclareMathOperator{\diag}{diag}
\newcommand{\ul}{\underline}

\usepackage{accents} % for the undertilde
\newcommand{\ut}{\undertilde}

@@

Before getting to the full SU($n$) Heisenberg model, we first revisit
the more common case of SU(2).  The SU(2) Heisenberg model consists of
$N$ two-level spins on a lattice, with spin-aligning interactions
\begin{align}
  H_0 = \sum_{\substack{\t{spin pairs}\\\set{i,j}}}
  h_{ij} \m s_i \c\m s_j,
  \label{eq:H_SU2}
\end{align}
where
$\m s_i\equiv\f12\p{\sigma_\x^{(i)},\sigma_\y^{(i)},\sigma_\z^{(i)}}$
is a spin vector defined with Pauli operators $\sigma_\alpha^{(i)}$
acting on the spin at lattice site $i$.  In the standard language of
spin physics, each term of this Hamiltonian essentially imposes an
energy penalty of $-h_{ij}$ to the formation of a spin 0 singlet state
$\ket{\up\dn}-\ket{\dn\up}$ between spins $i$ and $j$, favoring the
formation of spin triplet states with total spin 1.  Another
interpretation of the Hamiltonian in \eqref{eq:H_SU2} is illuminated
by the observation that the two-spin operator $\m s_i\c\m s_j$ is, up
to an overall constant, essentially the two-qubit SWAP operation
$\Pi_{ij}$.  We will refer to $\Pi_{ij}$ as {\it permutation
  operator}, because its action on the Hilbert space
$\H\equiv\bigotimes_{k=1}^N\H_k$ of $N$ spins with individual Hilbert
spaces $\H_k$ is to permute the tensor factors of $\H$ associated with
spins $i$ and $j$ (see Figure \ref{fig:sketch_spins}).  Up to identity
terms with no physical consequence, we then have
\begin{align}
  H_0 = \sum_{\substack{\t{spin pairs}\\\set{i,j}}} h_{ij} \Pi_{ij},
  &&
  \Pi_{ij} \equiv \sum_{\substack{\t{spin states}\\\mu,\nu}}
  s_{\mu\nu}^{(i)} s_{\nu\mu}^{(j)}
  = \sum_{\substack{\t{spin states}\\\mu,\nu}} \op{\mu\nu}{\nu\mu}_{ij},
  \label{eq:H_0}
\end{align}
where $s_{\mu\nu}^{(i)}\equiv\op{\mu}{\nu}_i$ is a spin-transition
operator for spin $i$, and $\mu,\nu\in\set{\up,\dn}$ index the states
of a single spin.  In addition to providing a nice interpretation of
the Hamiltonian $H_0$ as essentially swapping the states of spins $i$
and $j$ at a rate $h_{ij}$, the definition of the SU(2)-symmetric
Heisenberg model in in \eqref{eq:H_0} is straightforward to generalize
to SU($n$) with arbitrary $n$.  That is, rather than having
$\mu,\nu\in\set{\up,\dn}$ index two spin states, we simply let
$\mu,\nu\in\set{1,2,\cdots,n}$ index the orthogonal states (sometimes
called {\it colors} or {\it flavors}) of an $n$-level spin.

@@

If the coefficients $h_{ij}$ of the interaction Hamiltonian $H_0$ are
all negative, then the ground-state manifold $\M_0$ of $H_0$ consists
of spin-polarized, {\it permutationally symmetric} (PS) states that
are simultaneous $+1$ eigenstates of all permutation operators
$\Pi_{ij}$.  We will assume that all $h_{ij}<0$ throughout these
notes, but note that this restriction can be relaxed to the assumption
that the initial state of all spins is a permutationally-symmetric
state (e.g.~a spin-polarized state).  We will also assume that the PS
manifold $\M_0$ is gapped away from all orthogonal states by an
interaction energy $\Delta_{\t{gap}}>0$.  Power-law couplings of the
form $h_{ij}\sim-1/\abs{\m r_i-\m r_j}^\alpha$ on a $D$-dimensional
lattice, for example, always yield a non-vanishing spectral gap
$\Delta_{\t{gap}}$ for long-range interactions with $\alpha\le D$ (see
Appendix \ref{sec:spectral_gap}).  Here $\m r_i$ denotes the position
of lattice site $i$.

In the case of SU(2), the PS manifold $\M_0$ is precisely the {\it
  Dicke manifold} spanned by the PS states $\ket{m_\up,m_\dn}$ with a
definite (integer) number $m_\mu\ge0$ of spins occupying the
single-spin state $\mu\in\set{\up,\dn}$.  If there are $N$ spins
total, then Dicke states $\ket{m_\up,m_\dn}$ must have
$m_\up+m_\dn=N$.  In the case of multi-level spins, the PS manifold
$\M_0$ is similarly spanned by states
$\ket{m}=\ket{m_1,m_2,\cdots,m_n}$ with a definite occupation number
$m_\mu\ge0$ of the single-spin state $\mu$, with $\sum_\mu m_\mu=N$.
These states are essentially labeled by an assignment
$m=\p{m_1,m_2,\cdots,m_n}$ of $N$ (identical) spins to $n$ (distinct)
states.  We denote the set of all such assignments by $\A_n\p{N}$,
such that our standard basis for $\M_0$ is
$\set{\ket{m}:m\in\A_n\p{N}}$.  The dimension of $\M_0$, or
equivalently the ground-state degeneracy of $H_0$, is straightforward
to determine with a standard ``stars and bars'' calculation:
$\dim\p{\M_0}=\abs{\A_n\p{N}} = {N+n-1 \choose n-1} \sim N^{n-1}$.
The energy of any PS state $\ket\psi\in\M_0$ with respect to the
interaction Hamiltonian $H_0$ in \eqref{eq:H_0} is
\begin{align}
  E_0 = \sum_{\substack{\t{spin pairs}\\\set{i,j}}} h_{ij}
  = \sum_{i<j} h_{ij},
\end{align}
where $i,j\in\set{1,2,\cdots,N}$ index individual lattice sites.

@@

\vspace{3cm}

% todo: simplify, give examples at every step
% todo: consider special case before a general one
% todo: new symbol for X_w, e.g. O(X,w)

In order to induce non-trivial dynamics for initially PS states
$\ket\psi\in\M_0$, which are otherwise stationary eigenstates of the
interaction Hamiltonian $H_0$ in \eqref{eq:H_0}, we will add to the
Hamiltonian an operator $\hat\O$ of the general form
\begin{align}
  \hat\O \equiv \sum_{\p{X,\m w}\in\O} X_{\m w},
  &&
  X_{\m w} \equiv \f1{M!} \sum_{k\in\D_N\p{M}} w_k X_k,
  \label{eq:pert}
\end{align}
where $\hat\O$ is constructed from the set $\O\equiv\set{\p{X,\m w}}$
of $M$-spin operators $X$ together with dimension-$M$ (i.e.~$M$-index)
tensors $\m w$; $k=\p{k_1,k_2,\cdots,k_M}$ is a list of of $M$ spins
that $X_k$ acts on; $w_k$ is a scalar entry of $\m w$; and
\begin{align}
  \D_N\p{M} \equiv
  \set{ k \in \ZZ_N^M: \t{all entries of}~k~\t{are distinct} },
  \label{eq:off_diags}
\end{align}
is the strictly off-diagonal part of $\ZZ_N^M$.  The factor of $1/M!$
is included in \eqref{eq:pert} by convention to average over the $M!$
terms addressing any given set of spins.  To simplify language and
notation, we will assume throughout this work that $\hat\O$ is
homogeneous in the sense that $M$ is the same for all
$\p{X,\m w}\in\O$.  A generalization to the case of inhomogeneous
operators $\hat\O$ is straightforward.

Single-body ($M=1$) operators correspond to classical external fields,
e.g.~$s_\z$ for a magnetic field, and two-body ($M=2$) operators
correspond to two-body interactions that deviate from $H_0$, such as
$s_\z\otimes s_\z$ interactions in an XXZ model near the Heisenberg
critical point.  While we would be hard-pressed to find a physical
implementation of multi-body spin interactions with $M>2$, considering
this case will be necessary to construct low-lying eigenstates of
$H_0$, which will in turn enable simulating the dynamics induced by
non-perturbative additions $\hat\O$ to the SU($n$) Heisenberg model,
with operator norms $\norm*{\hat\O}$ comparable to the spectral gap
$\Delta_{\t{gap}}$ of $H_0$.

When the operator norm of $\hat\O$ is sufficiently small, namely
$\norm*{\hat\O}<\Delta_{\t{gap}}/2$, we can treat its effect on the PS
manifold $\M_0$ perturbatively.  The effective Hamiltonian
$H_{\t{eff}}$ induced by $\hat\O$ on the ground-state manifold $\M_0$
through second order in perturbation theory
is\cite{bravyi2011schrieffer, perlin2019effective}
\begin{align}
  H_{\t{eff}} = H_{\t{eff}}^{(1)} + H_{\t{eff}}^{(2)},
  &&
  H_{\t{eff}}^{(1)} = \P_0 \hat\O \P_0,
  &&
  H_{\t{eff}}^{(2)} = - \sum_{\Delta\ne0}
  \f1\Delta \P_0 \hat\O \P_\Delta \hat\O \P_0,
  \label{eq:H_eff}
\end{align}
where $\P_\Delta$ is a projector onto the eigenspace $\E_\Delta$ of
the interaction Hamiltonian $H_0$ with interaction energy $\Delta$
above that of PS manifold $\M_0$.  In particular, $\P_0$ is a
projector onto $\M_0$ itself, and $\E_0=\M_0$. The first order
effective Hamiltonian $H_{\t{eff}}^{(1)}$ simply projects the
perturbation $\hat\O$ onto the permutationally-symmetric manifold
$\M_0$, and takes the form
\begin{align}
  H_{\t{eff}}^{(1)} \EQPS \sum_{\p{X,\m w}\in\O} \bar w\,\col{X},
  \label{eq:H_eff_1}
\end{align}
where $\EQPS$ denotes equality under a restriction to $\M_0$; the
coefficient $\bar w$ is the average of all $w_k$; and the collective
operator $\col{X}$ is a sum of all $X_k$ averaged over permutations of
any particular choice of spins $k\in\D_N\p{M}$.  To write out the
collective operator $\col{X}$ explicitly, we define the set \begin{align}
  \C_N\p{M} \equiv \set{ k \subset \ZZ_N: \abs{k} = M },
  \label{eq:choices}
\end{align}
is the set of all subsets (``choices'') of $M$ elements from $\ZZ_N$,
and the set $\pi\p{M}$ of all permutations of $M$ elements, in terms
of which
\begin{align}
  \bar w \equiv \f1{\abs{\D_N\p{M}}}
  \sum_{k\in\D_N\p{M}} w_k,
  &&
  \col{X} \equiv \f1{\abs{\pi\p{M}}} \sum_{\sigma\in\pi\p{M}}
  \sum_{k\in\C_N\p{M}} X_{\sigma\p{k}},
\end{align}
where $\abs{\D_N\p{M}}=\prod_{j=0}^{M-1}\p{N-j}$ and
$\abs{\pi\p{M}}=M!$.  Here $k\in\C_N\p{M}$ defines which spins
$X_{\sigma\p{k}}$ acts on, while $\sigma\in\pi\p{M}$ determines which
tensor factor of $X$ each of these spins is associated with.  An
immediate consequence of \eqref{eq:H_eff_1} is the fact that
perturbing SU(2)-symmetric interactions by two-body $s_\z\otimes s_\z$
interactions yields the one-axis twisting (OAT)
Hamiltonian\cite{kitagawa1993squeezed, ma2011quantum}
$H_{\t{OAT}}=\chi S_\z^2$ at first order in perturbation theory, with
the OAT strength $\chi$ equal to the mean coefficient of the two-body
$s_\z\otimes s_\z$ terms.

The second order effective Hamiltonian $H_{\t{eff}}^{(2)}$ is not as
straightforward to simplify due to the presence of a projector
$\P_\Delta$ onto the manifold $\E_\Delta$ of states with excitation
energy $\Delta$.  For each term $X_{\m w}$ of $\hat\O$, this projector
essentially picks off the part of $X_{\m w}$ that is strictly
off-diagonal with respect to the ground- and excited-state manifolds
$\E_0$ and $\E_\Delta$.  We therefore need to decompose $X_{\m w}$
into components that generate states of definite excitation energy
when acting on PS states $\ket\psi\in\M_0$.  The SU($n$) symmetry of
$H_0$ enables such a decomposition of the form
\begin{align}
  H_0 X_{\m w}\ket\psi
  = \sum_\Delta \p{E_0+\Delta} X_{\m w_\Delta} \ket\psi,
\end{align}
where, thinking of the tensor $\m w$ as a $\abs{\D_N\p{M}}$-component
vector, the tensor $\m w_\Delta$ can be found found by diagonalizing a
matrix $\hat{\m h}$ of dimensions
$\abs{\D_N\p{M}}\times\abs{\D_N\p{M}}\sim N^M\times N^M$, and
projecting $\m w$ onto the eigenspace of $\hat{\m h}$ with eigenvalue
$\Delta$ (see Appendix \ref{sec:eigenstates}).  The matrix
$\hat{\m h}$ is determined entirely by the couplings couplings
$h_{pq}$ of $H_0$.  While the computational cost of diagonalizing
$\hat{\m h}$ is nominally $O\p{N^{3M}}$, spatial symmetries of $H_0$
and $X_{\m w}$ such as translational invariance or isotropy can
drastically reduce the complexity of this eigenvalue problem.  We
discuss this reduction in Appendix \ref{sec:symmetries}.

Equipped with the decomposition of $X_{\m w}$ into terms that generate
states of definite excitation energy, we can expand
\begin{align}
  H_{\t{eff}}^{(2)} = -\sum_{\Delta\ne0} \f1\Delta
  \sum_{\p{X,\m v},\p{Y,\m w}\in\O}
  \P_0 X_{\m v_\Delta} Y_{\m w_\Delta} \P_0.
\end{align}
Simplifying product $\P_0 X_{\m v_\Delta} Y_{\m w_\Delta} \P_0$
generally yields $r$-body operators for all positive integers
$r\le2M$.  We provide a general prescription for simplifying such
products in Appendix \ref{sec:operator_product}, and explicitly work
out single-body ($M=1$) and two-body ($M=2$) cases in Appendices
\ref{sec:single_pair_prod} and \ref{sec:two_pair_prod}.  If $X$ is a
single-body operator ($M=1$), then
\begin{align}
  H_{\t{eff}}^{(2)}
  = \sum_{\Delta\ne0} \sum_{\p{X,\m v},\p{Y,\m w}\in\O}
  \f{\cov\p{\m v_\Delta,\m w_\Delta}}{\p{N-1}\Delta}
  \p{\col{X}\,\col{Y} - N \col{XY}},
  \label{eq:H_eff_2_1}
\end{align}
where
\begin{align}
  \cov\p{\m v,\m w} \equiv
  \f1N \sum_{p\in\ZZ_N} \p{v_p-\bar v}\p{w_p-\bar w}
\end{align}
is the covariance between $\m v$ and $\m w$.  In the SU(2) case,
\eqref{eq:H_eff_2_1} implies that an inhomogeneous magnetic field
($X\propto s_\z$) yields an OAT Hamiltonian at second order
perturbation theory.

% briefly add expression for the third order Hamiltonian

@@

@@

@@

  H_0 \tilde X_{\m w} \ket\psi
  = \sum_{\substack{k\in\D_N\p{M}\\\set{p,q}\in\C_N\p{2}}}
  h_{pq} w_k \Pi_{pq} X_k \ket\psi,

@@

where to avoid factors of $M!$ in this section, we define
$\tilde X_{\m w}\equiv M! X_{\m w}$.  The sum in
\eqref{eq:diagnosis_start} has terms with $p,q\notin k$, terms with
$p,q\in k$, and terms with only one of $p$ or $q\in k$.  The
permutation operator $\Pi_{pq}$ acts trivially on $X_k\ket\psi$ when
$p,q\notin k$, allowing us to simplify

@@

  \sum_{\substack{k\in\D_N\p{M}\\\set{p,q}\in\C_N\p{2}\\p,q\notin k}}
  h_{pq} w_k \Pi_{pq} X_k \ket\psi
  = E_0 \tilde X_{\m w} \ket\psi

>>>>

%%% symbols, notations, etc.
\usepackage{physics,braket,bm,amssymb} % physics and math
\renewcommand{\t}{\text} % text in math mode
\newcommand{\f}[2]{\dfrac{#1}{#2}} % shorthand for fractions
\newcommand{\p}[1]{\left(#1\right)} % parenthesis
\renewcommand{\sp}[1]{\left[#1\right]} % square parenthesis
\newcommand{\spp}[1]{\sp{\sp{#1}}} % double square parenthesis
\renewcommand{\set}[1]{\left\{#1\right\}} % curly parenthesis
\newcommand{\bk}{\Braket} % shorthand for braket notation

\renewcommand{\c}{\cdot} % inner product
\newcommand{\m}{\bm} % bold symbol
\renewcommand{\v}{\vec} % arrow vector

@@

\renewcommand{\i}{\mathrm{i}\mkern1mu}
\renewcommand{\d}{\text{d}}
\newcommand{\x}{\text{x}}
\newcommand{\y}{\text{y}}
\newcommand{\z}{\text{z}}

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\G}{\mathcal{G}}
\renewcommand{\H}{\mathcal{H}}
\newcommand{\I}{\mathcal{I}}
\renewcommand{\L}{\mathcal{L}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\Q}{\mathcal{Q}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\T}{\mathcal{T}}

@@

\newcommand{\PS}{\text{PS}}
\newcommand{\EQPS}{=_{\text{PS}}}

\let\var\relax
\DeclareMathOperator{\var}{var}
\DeclareMathOperator{\cov}{cov}
\DeclareMathOperator{\diag}{diag}

\usepackage{accents} % for the undertilde
\newcommand{\ut}{\undertilde}
\newcommand{\ot}{\widetilde}
\newcommand{\col}{\underline}
\newcommand{\mean}{\overline}

@@

Before getting to the full SU($n$) Heisenberg model, we first revisit
the more common case of SU(2).  The SU(2) Heisenberg model consists of
$N$ two-level spins on a lattice, with spin-aligning interactions
\begin{align}
  H_0 = \sum_{i\ne j} h_{ij} \m s_i \c\m s_j,
  \label{eq:H_SU2}
\end{align}
where $i,j$ index individual spins; $h_{ij}$ are scalar coefficients
with $h_{ij}=h_{ji}$ and $h_{ii}=0$; and
$\m s_i\equiv\p{s_\x^{(i)},s_\y^{(i)},s_\z^{(i)}}$ is a vector of spin
operators $s_\alpha^{(i)}\equiv\sigma_\alpha^{(i)}/2$ defined by Pauli
operators $\sigma_\alpha^{(i)}$ acting on spin $i$.  In the standard
language of spin physics, each term of this Hamiltonian essentially
imposes an energy penalty of $-2h_{ij}$ to the formation of a spin 0
singlet state $\ket{\up\dn}-\ket{\dn\up}$ between spins $i$ and $j$,
favoring the formation of spin triplet states with total spin 1.
Another interpretation of the Hamiltonian in \eqref{eq:H_SU2} is
illuminated by the observation that the two-spin operator
$\m s_i\c\m s_j$ is, up to an overall constant, essentially the
two-qubit SWAP operation $\Pi_{ij}$.  We will refer to $\Pi_{ij}$ as
{\it permutation operator}, because its action on the Hilbert space
$\H\equiv\bigotimes_{k=1}^N\H_k$ of $N$ spins with individual Hilbert
spaces $\H_k$ is to permute the tensor factors of $\H$ associated with
spins $i$ and $j$ (see Figure \ref{fig:sketch_spins}).  Up to identity
terms with no physical consequence, we then have
\begin{align}
  H_0 = \sum_{i<j} h_{ij} \Pi_{ij},
  &&
  \Pi_{ij} \equiv \sum_{\mu,\nu} s_{\mu\nu}^{(i)} s_{\nu\mu}^{(j)}
  = \sum_{\mu,\nu} \op{\mu\nu}{\nu\mu}_{ij}
  \label{eq:H_0}
\end{align}
where $s_{\mu\nu}^{(i)}\equiv\op{\mu}{\nu}_i$ is a spin-transition
operator for spin $i$, and $\mu,\nu\in\set{\up,\dn}$ index the states
of a single spin.  In addition to providing a nice interpretation of
the Hamiltonian $H_0$ as essentially swapping the states of spins $i$
and $j$ at a rate $h_{ij}$, the definition of the SU(2)-symmetric
Heisenberg model in in \eqref{eq:H_0} is straightforward to generalize
to SU($n$) with arbitrary $n$.  That is, rather than having
$\mu,\nu\in\set{\up,\dn}$ index two spin states, we simply let
$\mu,\nu\in\set{1,2,\cdots,n}$ index the states (sometimes called {\it
  colors} or {\it flavors}) of an $n$-level spin.

@@

If the coefficients $h_{ij}$ of the interaction Hamiltonian $H_0$ are
all negative, then the ground-state manifold $\M_0$ of $H_0$ consists
of spin-polarized, {\it permutationally symmetric} (PS) states that
are simultaneous $+1$ eigenstates of all permutation operators
$\Pi_{ij}$.  We will assume that all $h_{ij}<0$ throughout these
notes, but note that this restriction can be relaxed to the assumption
that the initial state of all spins is a permutationally-symmetric
state (e.g.~a spin-polarized state).  We will also assume that the PS
manifold $\M_0$ is gapped away from all orthogonal states by an
interaction energy $\Delta_{\t{gap}}>0$.  Power-law couplings of the
form $h_{ij}\sim-1/\abs{i-j}^\alpha$ on a $D$-dimensional lattice, for
example, always yield a non-vanishing spectral gap $\Delta_{\t{gap}}$
for long-range interactions with $\alpha\le D$ (see Appendix
\ref{sec:spectral_gap}).

In the case of SU(2), the PS manifold $\M_0$ is precisely the {\it
  Dicke manifold} spanned by the PS states $\ket{m_\up,m_\dn}$ with a
definite (integer) number $m_\mu\ge0$ of spins occupying the
single-spin state $\mu\in\set{\up,\dn}$.  If there are $N$ spins
total, then Dicke states $\ket{m_\up,m_\dn}$ must have
$m_\up+m_\dn=N$.  In the case of multi-level spins, the PS manifold
$\M_0$ is similarly spanned by states
$\ket{m}=\ket{m_1,m_2,\cdots,m_n}$ with a definite occupation number
$m_\mu\ge0$ of the single-spin state $\mu$, and $\sum_\mu m_\mu=N$.
These states are essentially labeled by an assignment
$m=\p{m_1,m_2,\cdots,m_n}$ of $N$ (identical) spins to $n$ (distinct)
states.  We denote the set of all such assignments by $\A_n\p{N}$,
such that our standard basis for $\M_0$ is
$\set{\ket{m}:m\in\A_n\p{N}}$.  The dimension of $\M_0$, or
equivalently the ground-state degeneracy of $H_0$, is straightforward
to determine with a standard ``stars and bars'' calculation:
$\dim\p{\M_0}=\abs{\A_n\p{N}} = {N+n-1 \choose n-1} \sim N^{n-1}$.
The energy of any PS state $\ket\psi\in\M_0$ with respect to the
interaction Hamiltonian $H_0$ in \eqref{eq:H_0} is
\begin{align}
  E_0 = \sum_{i<j} h_{ij} = \f12 \sum_{i\ne j} h_{ij}.
\end{align}

@@

A common task, for example in state preparation protocols or quenched
dynamics experiments, is to first initialize in some trivial uniform
product state, $\ket\phi^{\otimes N}$, and then to let it evolve under
a (possibly time-dependent) Hamiltonian that induces non-trivial
dynamics.  The permutational symmetry of states such as
$\ket\phi^{\otimes N}\in\M_0$, however, makes them eigenstates of the
SU($n$)-symmetric interaction Hamiltonian $H_0$ in \eqref{eq:H_0}.  In
order to induce non-trivial dynamics, we therefore need to add new
terms to the Hamiltonian that break its SU($n$) symmetry.

As a concrete example, this sort of scenario can be used in the SU(2)
case to generate spin squeezed states that harness entanglement to
surpass classical limits on measurement
precision\cite{he2019engineering}\red{[cite XXZ squeezing paper]}.
Starting with a general (gapped) SU(2)-symmetric Hamiltonian of the
form in \eqref{eq:H_SU2} we can consider the addition of Ising
interactions
\begin{align}
  H_{\t{Ising}} = \sum_{i\ne j} J_{ij} s_\z^{(i)} s_\z^{(j)}.
  \label{eq:Ising}
\end{align}
If these interactions are sufficiently weak for perturbative
treatment, as in the case of the XXZ model near the Heisenberg
critical point, then the effective Hamiltonian induced by
$H_{\t{Ising}}$ on initially PS states $\ket\psi\in\M_0$ is the
one-axis twisting (OAT) Hamiltonian\cite{ma2011quantum}
\begin{align}
  H_{\t{OAT}} = \chi S_\z^2,
  \label{eq:OAT}
\end{align}
where the collective operator $S_\z\equiv\sum_{j=1}^Ns_\z^{(j)}$, and
the squeezing strength $\chi$ is equal to the mean Ising coupling
strength $J_{ij}$.  If Ising interactions are unavailable, the same
effect can be achieved by applying an inhomogeneous magnetic field,
\begin{align}
  H_{\t{B}} = \sum_i B_i s_\z^{(i)},
  \label{eq:magnetic}
\end{align}
which gives rise to an OAT Hamiltonian at second order in perturbation
theory, with a squeezing strength $\chi$ that is proportional to the
variance of the magnetic field strength $B_i$.

In order to go beyond first-order treatment of two-body interactions
such as \eqref{eq:Ising}, or beyond second order in the treatment of
single-body terms such as \eqref{eq:magnetic}, it will turn out to be
necessary to introduce $M$-body operators with $M>2$.  Generalizing
the above examples to the case of SU($n$) systems, we therefore
consider a general $M$-body operator of the form
\begin{align}
  \O\p{\m w,X} \equiv \sum_{k\in\D_N\p{M}} w_k X_k,
  \label{eq:multi_body_op}
\end{align}
where $\m w$ is a dimension-$M$ (i.e.~$M$-index) tensor of scalar
coefficients $w_k\equiv w_{k_1k_2\cdots k_M}$; $X$ is an $M$-spin
operator, e.g.~$s_\z\otimes s_\z$ in the case of Ising interactions
with $M=2$; $k\equiv\p{k_1,k_2,\cdots,k_M}$ is a list of the
individual spins $k_i\in\ZZ_N$ that the operator
$X_k\equiv X_{k_1k_2\cdots k_M}$ acts on; and
\begin{align}
  \D_N\p{M} \equiv
  \set{ k \in \ZZ_N^M : \t{all entries $k_i$ of $k$ are distinct } },
  \label{eq:off_diags}
\end{align}
is the strictly ``off-diagonal'' part of $\ZZ_N^M$, which is necessary
to identify for a consistent definition of $X_k$ as an $M$-body
operator.  In this notation, we can write the magnetic field and
Hamiltonians in \eqref{eq:magnetic} and \eqref{eq:Ising} as
$H_{\t{B}}=\O\p{\m B,s_\z}$ and
$H_{\t{Ising}}=\O\p{\m J,s_\z\otimes s_\z}$.

If the addition $\O\p{\m w,X}$ to the SU($n$)-symmetric Hamiltonian
$H_0$ in \eqref{eq:H_0} is sufficiently small, then we can treat the
effect of $\O\p{\m w,X}$ on the PS manifold $\M_0$ perturbatively.
The effective Hamiltonians $H_{\t{eff}}^{(1)}$ and $H_{\t{eff}}^{(2)}$
induced by $\O\p{\m w,X}$ on $\M_0$ at first and second order in
perturbation theory are\cite{bravyi2011schrieffer}
\begin{align}
  H_{\t{eff}}^{(1)} = \P_0 \O\p{\m w,X} \P_0,
  &&
  H_{\t{eff}}^{(2)} = - \sum_{\Delta\ne0}
  \f1\Delta \P_0 \O\p{\m w,X} \P_\Delta \O\p{\m w,X} \P_0,
  \label{eq:H_eff}
\end{align}
where $\P_\Delta$ is a projector onto the eigenspace $\E_\Delta$ of
the interaction Hamiltonian $H_0$ with interaction energy $\Delta$
above that of PS manifold $\M_0$.  In particular, $\P_0$ is a
projector onto $\M_0$ itself, and $\E_0=\M_0$.  The first order
effective Hamiltonian $H_{\t{eff}}^{(1)}$ simply projects
$\O\p{\m w,X}$ onto the PS manifold $\M_0$, and takes the form
\begin{align}
  H_{\t{eff}}^{(1)} \EQPS \mean w \, \col X,
  \label{eq:H_eff_1}
\end{align}
where $\EQPS$ denotes equality under a restriction to $\M_0$; the
coefficient $\mean w$ is the average of all $w_k$; and $\col X$ is a
collective version of $X$:
\begin{align}
  \mean w \equiv \f1{\abs{\D_N\p{M}}}
  \sum_{k\in\D_N\p{M}} w_k,
  &&
  \col X \equiv \sum_{k\in\D_N\p{M}} X_k,
\end{align}
with $\abs{\D_N\p{M}}=\prod_{j=0}^{M-1}\p{N-j}$.  In the case of a
magnetic field $s_\z$ or Ising interactions $s_\z\otimes s_\z$, for
example,
\begin{align}
  \col{s_\z} = \sum_i s_\z^{(i)} = S_\z,
  &&
  \col{s_\z\otimes s_\z}
  = \sum_{i\ne j} s_\z^{(i)} s_\z^{(j)}
  = S_\z^2 - \f14 N.
\end{align}
The second order effective Hamiltonian $H_{\t{eff}}^{(2)}$ in
\eqref{eq:H_eff} takes more work to simplify due to the presence of a
projector $\P_\Delta$ onto the manifold $\E_\Delta$ of states with
excitation energy $\Delta$.  This projector essentially picks off the
part of $\O\p{\m w,X}$ that is strictly off-diagonal with respect to
the ground- and excited-state manifolds $\E_0$ and $\E_\Delta$.  We
therefore need to decompose $\O\p{\m w,X}$ into components that
generate states of definite excitation energy when acting on PS states
$\ket\psi\in\M_0$.  The SU($n$) symmetry of $H_0$ enables such a
decomposition of the form
\begin{align}
  H_0 \O\p{\m w,X} \ket\psi
  = \sum_\Delta \p{E_0+\Delta} \O\p{\m w_\Delta,X} \ket\psi,
\end{align}
where, thinking of the tensor $\m w$ as a $\abs{\D_N\p{M}}$-component
vector, the tensor $\m w_\Delta$ can be found found by diagonalizing a
matrix $\hat{\m h}$ of dimensions
$\abs{\D_N\p{M}}\times\abs{\D_N\p{M}}\sim N^M\times N^M$, and
projecting $\m w$ onto the eigenspace of $\hat{\m h}$ with eigenvalue
$\Delta$ (see Appendix \ref{sec:eigenstates}).  The matrix
$\hat{\m h}$ is determined entirely by the couplings couplings
$h_{ij}$ of $H_0$.  While the computational cost of diagonalizing
$\hat{\m h}$ is nominally $O\p{N^{3M}}$, spatial symmetries of $H_0$
and $\O\p{\m w,X}$ such as translational invariance or isotropy can
drastically reduce the complexity of this eigenvalue problem, as we
discuss in Appendix \ref{sec:symmetries}.

Equipped with the decomposition of $\O\p{\m w,X}$ into terms
$\O\p{\m w_\Delta,X}$ that generate states of definite excitation
energy $\Delta$, we can expand
\begin{align}
  H_{\t{eff}}^{(2)} = -\sum_{\Delta\ne0} \f1\Delta
  \P_0 \O\p{\m w_\Delta,X}^2 \P_0.
\end{align}
Simplifying the product $\P_0 \O\p{\m w_\Delta,X}^2 \P_0$ generally
yields $r$-body operators for all positive integers $r\le2M$.  We
provide a general prescription for simplifying such products in
Appendix \ref{sec:operator_product}, and explicitly work out the
single-body ($M=1$) and two-body ($M=2$) cases in Appendices
\ref{sec:single_pair_prod} and \ref{sec:two_pair_prod}.  If $X$ is a
single-body operator, then
\begin{align}
  H_{\t{eff}}^{(2)}
  = \sum_{\Delta\ne0} \f{\var\p{\m w_\Delta}}{\p{N-1}\Delta}
  \p{\col{X}^2 - N \col{X^2}},
  \label{eq:H_eff_2_1}
\end{align}
where
\begin{align}
  \var\p{\m w} \equiv \f1{\abs{\D_N\p{M}}}
  \sum_{k\in\D_N\p{M}} \p{w_k - \mean w}^2
\end{align}
is the variance of $\m w$.  In the SU(2) case, this result implies
that an inhomogeneous magnetic field ($X=s_\z$) yields the OAT
Hamiltonian $H_{\t{OAT}}$ in \eqref{eq:OAT} at second order
perturbation theory.

\newpage

@@

\red{[note: this section is not yet finished or simplified]}

@@

\red{[look at \cite{rai1988correlated} for classifying asymmetric
  states]}

@@

  H_0 \O\p{\m w,X} \ket\psi
  = \sum_{\substack{k\in\D_N\p{M}\\\set{i,j}\in\C_N\p{2}}}
  h_{ij} w_k \Pi_{ij} X_k \ket\psi,

@@

@@

  \C_N\p{q} \equiv \set{ s \subset \ZZ_N : \abs{s} = q }
  \label{eq:choices}
is the set of all subsets (``choices'') of $q$ elements from $\ZZ_N$.
The sum in \eqref{eq:diagnosis_start} has terms with $i,j\notin k$,
terms with $i,j\in k$, and terms with only one of $i$ or $j\in k$.
The permutation operator $\Pi_{ij}$ acts trivially on $X_k\ket\psi$
